---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(table.glue)
library(tidyverse)
library(gt)
library(kableExtra)
library(gridExtra)

```

# rfvs-regression

**Citation for this work**: Nathaniel S O’Connell, Byron C Jaeger, Garrett S Bullock, Jaime Lynn Speiser, A comparison of random forest variable selection methods for regression modeling of continuous outcomes, *Briefings in Bioinformatics*, Volume 26, Issue 2, March 2025, bbaf096, https://doi.org/10.1093/bib/bbaf096


<!-- badges: start -->
<!-- badges: end -->

## Overview

The goal of rfvs-regression is to compare random forest variable selection techniques for continuous outcomes. We compare several methods available through various R packages and user defined functions from published paper appendices. The code for implementing each RF variable selection approach tested can be found in the function "rfvs()" and each of the methods assessed and compared are given in the following table:

<br/>

```{r}
tbl_meth <- read_csv("README_files/Table_of_RFVS_methods.csv", 
         show_col_types = FALSE)
tbl_meth <- as.data.frame(tbl_meth)

tbl_meth[,1:5] %>% kbl() %>% kableExtra::kable_styling()
```

<br/>

In this benchmarking study, we pulled datasets from *OpenML* and *modeldata* following the criteria and steps outlined below: 

# Datasets included

```{r}

# tbl <- read_csv("data/datasets_inclusion_chart.csv", 
#          show_col_types = FALSE) %>%  mutate(n=ifelse(n==76, 74, n)) %>% 
#          mutate(n= ifelse(n==61, 59, n)) %>%
#  gt(rowname_col = 'exclusion') %>% 
#  tab_footnote(locations = cells_stub(rows = 8), 
#               footnote = 'Many datasets are different versions of the same data')
# tbl
magick::image_read("README_files/figure-gfm/data_selection.png")

```


A total of 59 datasets met criteria and were used in this benchmarking study. Summary characteristics of these datasets are given in the figure below 

<center>
```{r}
magick::image_read("README_files/figure-gfm/Figure1.tiff")
```


# Benchmarking Study 

We used 20 replications of split sample validation (i.e., Monte-Carlo cross validation) for each dataset to evaluate RF variable selection methods.

1. First, a dataset was split into training (50%) and testing (50%) sets for all data sets with <2000 observations. For datasets with >2000 observations, 1000 observations were selected at random as the training set. This was done for computational time purposes; the largest datasets would take over 10 days for one replication/method. 

2. Second, each variable selection method was applied to the training data, and the variables selected by each method were saved.

 +     Note: For datasets with more than 150 predictors, a subset of 150 predictors were selected at random to serve as the variables for that replication


3. Third, a standard axis-based RF model using the R package *ranger* and an oblique RF using the package *aorsf* were fit on the training data set using variables selected from each method, and R-squared was recorded based on the test data for each replication, method, and dataset. 

4. Fourth, methods of variable selection were compared based on computation time, accuracy measured by R-squared, and percent variable reduction

*note*: If any missing values were present in the training or testing data, they were imputed prior to running variable selection methods using the mean and mode for continuous and categorical predictors, respectively, computed in the training data. 


## Primary Results Table

We provide the results in the table below for R-squared for downstream models fitted in Axis and Oblique RFs, variable percent reduction (higher % reduction implies more variables eliminated on average), and computation time (in seconds). 

```{r}
targets::tar_load(tab_main)
tab_main
```

## Primary Results Figure

We present the results of accuracy (x-axis), by percent reduction (y-axis), and time (size and color) in the figure below.  

```{r}
magick::image_read("README_files/figure-gfm/Figure7new.tiff")
```

We observe that for downstream Axis forests fitted in *ranger*, the methods of CARET (r packege: *caret*), Boruta (r package: *boruta*), Jiang, and aorsf-Permutation (r package: *aorsf*) perform optimally in terms of high-accuracy while preserving good parsimony (good variable percent reduction). Among them however, CARET is substantially slower on average, while Boruta, Jiang, and aorsf-Permutation are similar in accuracy and variable reduction, but substantially quicker in computation time.  

For downstream oblique forests fitted in *aorsf*, the methods aorsf-Menze and aorsf-Permutation (both found within the *aorsf* R package) perform best in terms of computation time and accuracy, while preserving good variable reduction. 

## Sensitivity to incomplete replication

We note that in several dataset replications, at least one method failed to select a single variable in variable selection. We performed a sensitivity analysis by assessing results only in replications where all methods selected at least one variable (10,192 replications out of 16,520 possible). 

```{r}
magick::image_read("README_files/figure-gfm/Figure8new.tiff")
```

## Comparison of Axis vs Oblique fitted downstream RFs

Last but not least, we compare downstream fitted Axis RFs to Oblique RFs. 

```{r}
magick::image_read("README_files/figure-gfm/Figure6.tiff")
```

We find that in terms of median accuracy, downstream oblique fitted forests generally perform better than downstream Axis forests, particularly among the top performing methods of variable selection. However, this is likely due to Oblique forest handling sparse data (N/P < 10) better than Axis forests. When we assess each stratified by N/P < or >= 10, we find that Axis and Oblique forests are much closer (with Oblique forests still performing slightly better among the top performing VS methods). 

```{r}
magick::image_read("README_files/figure-gfm/Figure9.tiff")
```
